---
title: R for Data Science
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message=F, warning=F)
```

```{r render-github-doc, eval=F, echo=F}
rmarkdown::render(input = "_Rmd/2023-07-26-r-for-data-science.Rmd", 
                  output_format = "github_document", 
                  output_dir = "_posts", 
                  output_options = list(html_preview = F))
```

Why I Now Prefer R for Data Science

In the growing landscape of programming languages used for data science tools, R stands out as my top choice. While I have experience with other languages such as SAS and Python, I find that R most aligns with my preferences. A few key reasons underpin my inclination:

1. Syntax and Structure: R's syntax is intuitive and well-structured, offering a user-friendly experience.
2. RStudio: This integrated development environment (IDE) not only makes R coding more efficient but also enhances the overall user experience.
3. Statistical Capabilities: R's roots in statistics are evident in its comprehensive array of statistical packages and functions, some of which are unmatched in other languages.
4. Versatility: Contrary to some beliefs, R matches, if not surpasses, the capabilities of Python and SAS in its data science capabilities

## Highlights from the NCSU *Data Science for Statisticians* Course

This course has been instrumental in deepening my understanding of machine learning within the R ecosystem. While I had some prior machine learning experience with Python, navigating the same functionality with R was enlightening. I was particularly impressed with the `caret` package, which I now consider to the the `scikit-learn` of the R language. 

Although had prior (self-taught) knowledge of some ML principles, this course allowed me to delve deeper into specific algorithms like linear regression, decision trees, bagging, boosting, and more, solidifying my foundational knowledge.

## Looking Ahead

In upcoming projects (either for work or for school), I intend to increasingly use R - especially for data science. Additionally, my interest in spatial statistics aligns well with R's capabilities in handling GIS and spatial data analysis techniques. 

As the field of data science continues to evolve, I believe that R will remain at the forefront of my toolkit, proving indispensable in both current and future endeavors.

