---
title: R you ready for this? My favorite features of R
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message=F, warning=F)
```


```{r render-github-doc, eval=F, echo=F}
rmarkdown::render(input = "_Rmd/2023-06-26-r-you-ready-for-this.Rmd", 
                  output_format = "github_document", 
                  output_dir = "_posts", 
                  output_options = list(html_preview = F))
```

## What's the best part about the R language? Well... It's hard to choose

I have never been good at "picking favorites." I absolutely despise get-to-know-you questions like "What is your favorite food?" Or "Where is your dream vacation?" I am the type of person who appreciates variety, practicality, and novelty. These are the very qualities that make the R language so appealing to me. For this post, (in the interest of brevity), I will focus in on novelty and highlight a selection of unique features exclusive to the R language.

### Formula Syntax

While other languages might have similar ways to specify models, the specific syntax and flexibility of R's formula interface is quite unique. It's one of the features that makes R particularly well-suited for statistical modeling and data analysis.

For example, a simple linear regression:

```{r formula-example}
# Load the built-in 'mtcars' dataset
data(mtcars)

# Fit a linear regression model using the formula syntax
model <- lm(mpg ~ cyl + hp, data = mtcars)

# Print the model summary
summary(model)
```

In this example, `mpg ~ cyl + hp` is a formula that specifies a model where the miles per gallon (mpg) is explained by the number of cylinders (cyl) and horsepower (hp).

### Piping

Although the concept itself is not unique (other languages such as Python or JavaScript have "chaining"), piping is an elegant way of keeping R code organized, interpretable, and efficient. The pipe operator `%>%` (which more recently has been updated to alternatively be `|>`) is simply placed between functions, linking them into a continuous process. For example:

```{r piping-example}
# Load the necessary library
library(magrittr) 
# Can also be loaded through dplyr, which requires magrittr

# Use piping to filter and summarize the mtcars dataset
mtcars %>%
  subset(cyl == 6) %>%
  summarise(mean_mpg = mean(mpg), mean_hp = mean(hp))
```

### Built-In Data Frame Data Structure

In R, data frames are built into the base language, and many of R's built-in functions and packages are designed to work seamlessly with them. This makes data frames in R particularly easy (and efficient) to work with. Although the concept itself is not unique to R (e.g., the Python language is known for its pandas library), the implementation of a data frame type within the base language sets R apart as a powerful language for data analytics and data science. 

The built-in `mtcars` R dataset, as used in the prior two examples, is an example of an R data frame (see the data structure below):

```{r data-frame-example}
str(mtcars)
```


